\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2023-03-08 response.md

Dear Reviewers,

Many thanks for your considered feedback and overall positive response.  Below we respond to the main questions raised in the reviews 19A and 10C (as no questions were raised in #19B and #19D).  We are confident that we can address all issues with a minor revision of our manuscript.

# Review #19A
We agree with the reviewer's constructive comments. To further clarify the discussion, we answered some questions raised in the review.

**Question**: "Through three studies involving real programmers" -- Are some programmers not real?

**Answer**: The intention was to differentiate mixed experienced programmers from university students. With that said, we agree with the reviewer that our wording could be more precise.

**Question**: Tasks were less than 1 minute in length? Seems quite short.

**Answer**: The mean task time for user study 1a is less than 1 minute. And admittedly, the user study 1a tasks were relatively trivial, as pointed out in the RQ1 discussion on page 7. The mean time in task 1b and study 2 are substantially longer.

**Question**: What is "/chameleon" mentioned on page 8? Looks like a broken LaTeX macro maybe.

**Answer**: Yes, it's a broken LaTex macro. We will conduct more thorough proofreading should the paper be accepted.

----

# Review #19C

## Questions for authors' response:

**Question**: How did the authors recognize a type error as "fixed"?

**Answer**: A type error is treated as "fixed" when the first time the program type check and some functional tests succeed. Functional tests are in place to prevent participants from "fixing" type errors by deleting offending code, as the reviewer noted. This information is provided in IV-A, data collection subsection (p. 7).

**Question**: In RQ1, why is a paired t-test possible?

**Answer**: The paired t-test is possible because the effect was measured on the same participant receiving both treatments (GHC compiler error message and ChameleonIDE). In section IV-A (p. 7), we noted the with-in-subject design of our human studies and provided justification. To further clarify,  In studies 1a and 1b, each participant received 8 tasks, 4 using the ChameleonIDE and 4 using GHC. The treatment alternated throughout the study (e.g., ChameleonIDE for tasks 1,3,5,7, GHC for tasks 2,4,6,8, or the same configuration with the order of tools reversed). The two configurations were counterbalanced within the population. We agree with the reviewer this information should more explicitly stated in the manuscript.

## Other questions raised in the review
We agree with the constructive comments made by the reviewer.
Some of the review's questions were answered here to help clarify the discussion.

**Question**: What kind of tasks are used in Study 1b and Study 2?
Only the sentence "in study 1b we introduced more difficult challenges" is the explanation. I could see source files on HotCRP but I could not understand whether they are more practical cases or not. The paper should show a summary of type errors and programs.

**Answer**: The tasks given in the supplementary material were the same tasks used in the studies. We agree with the reviewer the paper would benefit from a concise summary of the task programs or a synopsis of the supporting material.

**Question**: How did the authors deal with participants who switched the IDE modes during the tasks in Fig.17? Please clarify the number of data points. In addition, please explain the risk of biased results, as participants were able to switch to advanced modes.

**Answer**: The task time shown in  Fig.17 is not grouped by each mode of ChameleonIDE but rather by different levels of interactivity of participants regardless of their mode assignments. The justification was given in RQ 2.1 discussion on page 8. For instance, if a participant voluntarily switches modes multiple times, even to less advanced modes, during resolving type error, they are placed in the high interactivity group. Conversely, participants in advanced mode can still be placed in the minimal interaction group if they lack interaction with the tool.

**Question**: The availability of the tool is not described.

**Answer**: ChameleonIDE is open source and is available in web and desktop versions. The information is provided at the end of paragraph 4 on page 1.

### 2023-03-08 sections/evaluation.tex
Camera Ready

\subsubsection*{\textbf{Data Collection}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
After completing all tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a ``poor man's video recording" to identify usability issues in the study and to recognize general patterns. 
%The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
%One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim=0mm 12mm 0mm 35mm,clip]{images/user-study-1a.pdf}
    \caption{Study 1a task completion time (secs.) with 95\% confidence interval.}
    \label{fig:analysis-1a}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-1b.pdf}
    \caption{Study 1b task completion time (secs.) with 95\% confidence interval.}
    \label{fig:analysis-1b}
\end{figure}

@@

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse. Tukey’s HSD Test for multiple comparisons found that the completion time was significantly different between the minimal interaction group and the high interaction group ($p \le 0.001$, 95\% C.I. = [18.26, 31.41]), and between the minimal interaction group and the low interaction group ($p \le 0.001$, 95\% C.I. = [11.96, 26.67]). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by /chameleon{}. This reduced the apparent benefit of /chameleon{} in these tasks. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

@@

One threat to the validity of the evaluation is the number of participants. Although for each study we received hundreds of online participants, the studies suffered from a high abandonment rate (especially study 1b). This was expected: the programming challenges are difficult, and our volunteer participants are unremunerated. 
Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one. This creates variation in familiarity. We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues. To mitigate this, we conducted cognitive walkthroughs and sandbox pilots before running each study.

Future evaluation would benefit from using more realistic tasks. In studies 1b and 2, we used more challenging tasks compared to study 1a. However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase. It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include more confusing abstractions, like Monads, Monad transformers, and Lenses.

% The online study limited the amount of qualitative data we could gather from the users. Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods. Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.

>>>>

\subsubsection*{\textbf{Data Collection}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. Participants are able to skip a task if they are stuck. 
% The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
After completing all tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This allows us to identify usability issues in the study and to recognize general patterns. 
%The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
%One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim=15mm 12mm 15mm 35mm,clip]{images/user-study-1a.pdf}
    \caption{Study 1a task completion time (secs.) with 95\% confidence interval.}
    \label{fig:analysis-1a}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim=12mm 15mm 12mm 35mm,clip]{images/user-study-1b.pdf}
    \caption{Study 1b task completion time (secs.) with 95\% confidence interval.}
    \label{fig:analysis-1b}
\end{figure}

@@

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse. Tukey’s HSD Test for multiple comparisons found that the completion time was significantly different between the minimal interaction group and the high interaction group ($p \le 0.001$, 95\% C.I. = [18.26, 31.41]), and between the minimal interaction group and the low interaction group ($p \le 0.001$, 95\% C.I. = [11.96, 26.67]). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by \chameleon{}. This reduced the apparent benefit of \chameleon{} in these tasks. On the other hand, task 9 is the lengthiest task of all. It also involves deeply nested type definations that are harder to follow in mind.

@@

One threat to the validity of the evaluation is the number of participants. Although for each study we received hundreds of online participants, the studies suffered from a high abandonment rate (especially study 1b). This was expected: the programming challenges are difficult, and our volunteer participants are unremunerated. 
Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one. This creates variation in familiarity. We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
Conducting studies remotely and unsupervised left us no means to intervene when users encounter usability issues. To mitigate this, we conducted cognitive walkthroughs and sandbox pilots before running each study.

Future evaluation would benefit from using more realistic tasks. The tasks in our human studies do not get as complex as professional Haskell programmers may face in a typical production codebase. It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include more confusing abstractions, like Monads, Monad transformers, and Lenses.

### 2023-03-08 sections/introduction.tex
Camera Ready

This paper introduces \chameleon{}, an interactive type debugging tool for Haskell. It can visualize the relevant context of a type error: where it happens or could have happened and which parts of the code cause it. In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity. The most noticeable features are the type compare tool (Section \ref{sub:type-compare}), the candidate expression card (Section \ref{sub:candidate-expression}), and the deduction step (Section \ref{sub:deduction-steps}). These features are integrated into a debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available in web~\cite{anonymous_chameleon_2022} and desktop versions.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

@@

Our experiments showed that programmers using \chameleon{} fix type errors faster than with traditional text-based error messages. This difference is more significant when solving harder tasks. Further, programmers who actively use \chameleon{} interactive features fix type errors faster than simply reading the type error output.

>>>>

This paper introduces \chameleon{}, an interactive type debugging tool for Haskell. It can visualize the relevant context of a type error: where it happens or could have happened and which parts of the code cause it. In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity. The most noticeable features are the type compare tool (Section \ref{sub:type-compare}), the candidate expression card (Section \ref{sub:candidate-expression}), and the deduction step (Section \ref{sub:deduction-steps}). These features are integrated into a debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available at ~\cite{anonymous_chameleon_2022}.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

@@

Our experiments showed that programmers using \chameleon{} fix type errors faster than with traditional text-based error messages. This difference is more significant when solving harder tasks. Further, programmers who actively use \chameleon{} interactive features fix type errors faster than simply reading the type error output. Although \chameleon{} is designed to work with
the Haskell language, we plan to extend the underlying ideas to work with other strongly typed languages.

### 2023-03-08 sections/related-work.tex
Camera Ready

\subsection{Interactive Debugging}
The original Chameleon provides a certain level of ``interactive'' debugging through a command-line interface~\cite{stuckey_interactive_2003}, allowing the programmer to query different parts of the program. A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT (Counter-factual typing)~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} inference. Both tools allow programmers to change their assumptions about which expression is ill-typed. While both tools show a strong influence on theoretical development, they were not evaluated in realistic studies with human participants. In addition, command line-based debugging tools suffer the same limitation as other line-based tools, namely, limited user interaction, limited graphic capability (ANSI text), and the requirement of remembering the command language.

A GUI provides programming tools with immediate feedback on changes and the ability to show complex visualizations. Many tools aim to improve the debugging experience. We list two. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (called ``strategies'' by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like ``why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. These debugging tools are important motivations for developing \chameleon{}. However, they focus on different aspects of the debugging process. Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in development assistance supported by type holes.

>>>>

\subsection{Interactive Debugging}
% The original Chameleon provides a certain level of ``interactive'' debugging through a command-line interface~\cite{stuckey_interactive_2003}, allowing the programmer to query different parts of the program. A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT (Counter-factual typing)~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} inference. Both tools allow programmers to change their assumptions about which expression is ill-typed. While both tools show a strong influence on theoretical development, they were not evaluated in realistic studies with human participants. In addition, command line-based debugging tools suffer the same limitation as other line-based tools, namely, limited user interaction, limited graphic capability (ANSI text), and the requirement of remembering the command language.

Modern programming tools can offer alternative methods of code authoring, display real-time feedback and reveal complex programming contexts through visualizations. Many tools aim to improve the debugging experience using such capabilities. We list two. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (called ``strategies'' by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like ``why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. These debugging tools are important motivations for developing \chameleon{}. However, they focus on different aspects of the debugging process. Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in development assistance supported by type holes.

### 2023-03-08 sections/walkthrough.tex
Camera Ready

For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
the options provided by \chameleon{}, by switching to advanced
mode, she has access to the debugging steps.

First, Maxine clicks on step 5 (Fig. \ref{fig:advanced-mode-step5}) and verifies
that the two occurrences of \texttt{evens} are supposed to be identical, and the
second use means \texttt{evens} is a list of integers. Second, she
clicks on step 6 (Fig. \ref{fig:advanced-mode-step6}) and verifies that
\texttt{evens} should be the same type as the declaration on the right-hand
side.

>>>>

% For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
% the options provided by \chameleon{}, by switching to advanced
% mode, she has access to the debugging steps.

To illustrate the deduction steps with the task shown in section \ref{sub:balanced},  first, Maxine clicks on step 5 (Fig. \ref{fig:advanced-mode-step5}) and verifies
that the two occurrences of \texttt{evens} are supposed to be identical, and the
second use means \texttt{evens} is a list of integers. Second, she
clicks on step 6 (Fig. \ref{fig:advanced-mode-step6}) and verifies that
\texttt{evens} should be the same type as the declaration on the right-hand
side.
